Data Specialization for a Staged Lambda Calculus

### 2015-02-28 paper/abstract.tex

Staged programming languages assign a stage to each expression and
evaluate each expression in its assigned stage.
%For example, in a
%two-staged program, some expressions are evaluated in the first stage
%and others in the second stage.
%Programming languages, evaluation
%strategies, and type systems for staged programs where stages roughly
%correspond to the arrival order of inputs have been studied
%extensively in the areas of partial evaluation and meta programming.
While staged programming has been studied extensively in the areas of
partial evaluation and meta programming, in this paper we present an
algorithm for statically {\em splitting} programs witten in a
two-stage typed lambda calculus (with a $\fut$ modality denoting
computation in a later stage) into multi-pass programs that are
structured as two dependent, unstaged programs and are evaluated using
two distinct execution passes.  While splitting, also called pass
separation or data specialization, has been studied in the past, prior
techniques were either manual or applicable only to simple programs
that lack important features such as recursion and first-class
functions.

Since our work builds on a lambda calculus, our splitting algorithm
enables the programmer to write staged programs by using powerful
programming abstractions in a composable and modular fashion. Our
splitting transformation maps such programs into multi-pass programs
that can run significantly more efficiently by assigning frequent
computations to earlier passes so that they can be reused (cheaply)
later.  Our experiments based on a prototype implementation show that
our splitting algorithm is able perform highly non-trivial algorithmic
transformations, sometimes automatically improving efficiency
asymptotically by nearly a linear factor.  Since splitting
transformation takes place at compile time, these benefits are
achieved without significant run-time penalty.

>>>>

Staged programming languages assign a stage to each expression and
evaluate each expression in its assigned stage.
%For example, in a
%two-staged program, some expressions are evaluated in the first stage
%and others in the second stage.
%Programming languages, evaluation
%strategies, and type systems for staged programs where stages roughly
%correspond to the arrival order of inputs have been studied
%extensively in the areas of partial evaluation and meta programming.
While staged programming has been studied extensively in the areas of
partial evaluation and meta programming, in this paper we present an
algorithm for statically {\em splitting} programs written in a
two-stage typed lambda calculus (with a $\fut$ modality denoting
computation in a later stage) into multi-pass programs that are
structured as two dependent, unstaged programs and are evaluated using
two distinct execution passes.  While splitting has been studied in the past
under the names \emph{pass separation} and \emph{data specialization},
such techniques were either manual or applicable only to simple programs
that lack important features such as recursion and first-class
functions.

Since our work builds on a lambda calculus, our splitting algorithm
enables the programmer to write staged programs by using powerful
programming abstractions in a composable and modular fashion. Our
splitting transformation maps such programs into multi-pass programs
that can run significantly more efficiently by assigning frequent
computations to earlier passes so that they can be reused (cheaply)
later.  Our experiments based on a prototype implementation show that
our splitting algorithm is able to perform highly non-trivial algorithmic
transformations, sometimes automatically improving efficiency
asymptotically by nearly a linear factor.  Since the splitting
transformation takes place at compile time, these benefits are
achieved without significant run-time penalty.

### 2015-02-28 paper/relatedwork.tex

Frequency reduction and precomputation are common techniques for both
designing algorithms and performing compiler
optimizations~\cite{JS86-staging}.  The idea behind frequency
reduction is to identify computations that are performed multiple
times and pull them ahead so that they can be performed once and used
later as needed.  Dynamic programming, loop hoisting, and pass
separation (presented here) are all examples of frequency reduction
techniques.
%
The idea behind precomputation is to identify computations that can be
performed earlier (e.g., at compile time) if their inputs are
available statically and perform them at that earlier time. Dynamic
algorithms, partial evaluation, and incremental computation are all
examples of precomputation techniques.

\paragraph{Precomputation techniques.}
Perhaps one of the most well studied example of precomputation is
partial evaluation.  Partial evaluation distinguishes between static
(compile-time) and dynamic (run-time) stages. Given a program and values 
for all static inputs (inputs available at compilation
time), partial evaluation generates a specialized program by
performing computations that depend only on the static
inputs~\cite{jones96}.  The roots of partial evaluation go back to
Kleene, who in 1952 proved that a multivariate program (function)
$f(x_1, \ldots, x_{n+m})$ can be specialized for any given $a_1 \ldots
a_m$ to obtain another program $g(x_1, \ldots, x_n)$ such that $f(a_1,
\ldots, a_m, y_1, \ldots, y_n) = g(y_1, \ldots, y_n)$~\cite{Kleene52}.
Kleene did not intend to improve efficiency, however.  Lombardi
appears to be the first person to use the term in his work on
incremental computation~\cite{Lombardi67}.  There has since been 
an enormous body of work on partial evaluation, and we refer the reader to the
book by Jones, Gommard, and Sestoft for a comprehensive discussion of
work until the early 90's~\cite{JGS93}.

Early approaches to partial evaluation can be viewed as operating in
two stages: binding time analysis and program specialization.  For a
multivariate program with clearly marked static and dynamic arguments,
binding-time analysis identifies all the parts of the program that can
be computed by the knowledge of static arguments. Using this
information and the values of static arguments, program
specialization specializes the original program to a partially evaluated
one that operates on many different dynamic arguments.  This
approach has been applied to construct partial evaluators for a number
of languages such as Scheme~\cite{OB91-Similix,Consel88-Schism},
considering for the most part first-order subsets.

Experience with binding time analysis showed that it can be difficult
to control, leading to programs whose performance was difficult to
predict. This led to investigations based on type systems for making
the stage of computations explicit in programs~\cite{GJ91-lambda,NN92-twolevel}
and writing ``meta programs'' that, when evaluated at a stage, yield
a new program to be evaluated at the next stage.
%J{\o}rring and
%Scherlis describe the idea of meta programming as one of the three possible approaches to
%taking advantage of staging~\cite{JS86-staging}.
Davies~\cite{davies96} presented a
logical construction of binding-time type systems by deriving a type
system via the Curry-Howard isomorphism applied to temporal logic.
Davies and Pfenning proposed a new type system for staged computation
based on a particular fragment of modal logic~\cite{DP01-modal}. The
work on MetaML extended type-based techniques to a full-scale language
by developing statically typed programming language based on ML that
enables the programmer to express programs with multiple
stages~\cite{Taha97,taha-thesis-99}.  MetaML's type system is similar
to Davies~\cite{davies96} but extends it in several important ways.
Nanevksi and Pfenning further extended the these techniques by
allowing free variables to occur within staged
computations~\cite{NP05-nn}.

The type-system of \lang\ is closely related to this later line of
work on metaprogramming and staged computation.  The specific
extension to the typed lambda calculus that we use here is based on
the circle-modality of Davies~\cite{DP01-modal}.  Our types differ in
small ways to account for the specific evaluation model that we
propose, however the key difference between our work and this prior work is
that we instead focus on the problem of pass separation.
%we developed an algorithm
%for splitting terms in the proposed staged language.
%the staging or the splitting algorithm, which given a two-staged
%program splits it into two stages that can be executed sequentially,
%passing information from the earlier to the latter stage by
%construction.
%
%Unlike partial evaluation, splitting requires no knowledge of program
%argument values, only their staging.
%Using just the staging
%information, the splitting algorithm can generate programs that can
%perform non-trivial transformations on the code automatically.  As our
%examples illustrate such staging transformations can improve the
%efficiency of the source program, sometimes in asymptotically
%significant ways.

Another class of pre-computation techniques is incremental
computation, where a program can efficiently respond to small changes
to its input by only recomputing parts of the computation affected
by the
changes~\cite{DemersReTe81,PughTe89,RamalingamRe93,AcarBlBlHaTa09}.
Like other pre-computation techniques, incremental computation can be
thought of as a staged model, where changes in inputs come over a
period of time in stages.
%
However, incremental computation does not requiring fixing any of its
inputs and allows, in the general case, for all inputs to
change. Thus, its benefits are dependent the input changes made and
techniques typically utilize significant run-time (dynamic)
information.  For example, while we can apply incremental computation
to our quickselect example (\ref{sec:overview}), doing so would not
realize the same bounds as splitting since incremental computation
techniques would unfold the quickselect function based on the demanded
ranks, which can require linear time (as opposed to logarithmic) at
each stage.  Such an implementation would also maintain sophisticated
data structures dynamically at run time, the benefits of which depend
highly on input changes applied.

\paragraph{Frequency-reduction via pass separation.}
Perhaps the closest work to ours is the algorithm for ``data
specialization'' given by Knoblock and Ruf~\cite{knoblock96}, which
also seeks to statically split an explicitly staged program into two
stages.  However, they only consider a simple first-order language and
straight-line programs. Their work also treats all computations
guarded by second-stage conditionals as second-stage computations,
which would prevent optimization (via splitting) of programs such as
quickselect.

Automatic pass separation has also been a topic of interest in
programming systems for computer graphics, where, to achieve
high-performance, programs are typically manually separated into
components by frequency of execution (Separation into a pre-determined
set of per-frequency modules is \emph{required} by the execution model
of modern graphics architectures.)  The software engineering
challenges modifying multiple per-stage programs have led to
suggestions of writing graphics programs in an explicitly staged
programming language\,\cite{Proudfoot:2001,Foley:2011,He:2014}, and
deferring the task of pass separation to the compiler. However, all
prior efforts, like that of Knoblock and Ruf\,\cite{knoblock96} have
been limited to simple, imperative languages. We are not aware of
prior work the presents a splitting algorithm in the context of
higher-order functional languages.

>>>>

Frequency reduction and precomputation are common techniques for both
designing algorithms and performing compiler
optimizations~\cite{JS86-staging}.
%
The idea behind precomputation is to identify computations that can be
performed earlier (e.g., at compile time) if their inputs are
available statically and perform them at that earlier time. Dynamic
algorithms, partial evaluation, and incremental computation are all
examples of precomputation techniques.
%
The idea behind frequency
reduction is to identify computations that are performed multiple
times and pull them ahead so that they can be performed once and used
later as needed.  Dynamic programming, loop hoisting, and pass
separation (presented here) are all examples of frequency reduction
techniques.

\paragraph{Precomputation techniques.}
Perhaps one of the most studied examples of precomputation is
partial evaluation.  Partial evaluation distinguishes between static
(compile-time) and dynamic (runtime) stages. Given a program and values 
for all static inputs, partial evaluation generates a specialized program by
performing computations that depend only on the static
inputs~\cite{jones96}.  The roots of partial evaluation go back to
Kleene, who in 1952 proved that a multivariate program (function)
$f(x_1, \ldots, x_{n+m})$ can be specialized for any given $a_1 \ldots
a_m$ to obtain another program $g(x_1, \ldots, x_n)$ such that $f(a_1,
\ldots, a_m, y_1, \ldots, y_n) = g(y_1, \ldots, y_n)$~\cite{Kleene52}.
Kleene did not intend to improve efficiency, however.  Lombardi
appears to be the first person to use the term in his work on
incremental computation~\cite{Lombardi67}.  There has since been 
an enormous body of work on partial evaluation, and we refer the reader to the
book by Jones, Gommard, and Sestoft for a comprehensive discussion of
work until the early 90's~\cite{JGS93}.

Early approaches to partial evaluation can be viewed as operating in
two stages: binding time analysis and program specialization.  For a
multivariate program with clearly marked static and dynamic arguments,
binding-time analysis identifies all the parts of the program that can
be computed by the knowledge of static arguments. Using this
information and the values of static arguments, program
specialization specializes the original program to a partially-evaluated
one that operates on many different dynamic arguments.  This
approach has been applied to construct partial evaluators for a number
of languages such as Scheme~\cite{OB91-Similix,Consel88-Schism}, 
although for the most part restricted to first-order subsets.

Experience with binding time analysis showed that it can be difficult
to control, leading to programs whose performance was difficult to
predict. This led to investigations based on type systems for making
the stage of computations explicit in programs~\cite{GJ91-lambda,NN92-twolevel}
and writing ``metaprograms'' that, when evaluated at a stage, yield
a new program to be evaluated at the next stage.
%J{\o}rring and
%Scherlis describe the idea of meta programming as one of the three possible approaches to
%taking advantage of staging~\cite{JS86-staging}.
Davies~\cite{davies96} presented a
logical construction of binding-time type systems by deriving a type
system via the Curry-Howard correspondence applied to temporal logic.
Davies and Pfenning proposed a new type system for staged computation
based on a particular fragment of modal logic~\cite{DP01-modal}. The
work on MetaML extended type-based techniques to a full-scale language
by developing a statically typed programming language based on ML that
enables the programmer to express programs with multiple
stages~\cite{Taha97,taha-thesis-99}.  MetaML's type system is similar
to Davies~\cite{davies96} but extends it in several important ways.
Nanevksi and Pfenning further extended the these techniques by
allowing free variables to occur within staged
computations~\cite{NP05-nn}.

The type-system of \lang\ is closely related to this later line of
work on metaprogramming and staged computation.  The specific
extension to the typed lambda calculus that we use here is based on
the $\fut$ modality of Davies~\cite{DP01-modal}.  Our types differ in
small ways to account for the specific evaluation model that we
propose; however, the key difference between our work and this prior work is
that we instead focus on the problem of pass separation.
%we developed an algorithm
%for splitting terms in the proposed staged language.
%the staging or the splitting algorithm, which given a two-staged
%program splits it into two stages that can be executed sequentially,
%passing information from the earlier to the latter stage by
%construction.
%
%Unlike partial evaluation, splitting requires no knowledge of program
%argument values, only their staging.
%Using just the staging
%information, the splitting algorithm can generate programs that can
%perform non-trivial transformations on the code automatically.  As our
%examples illustrate such staging transformations can improve the
%efficiency of the source program, sometimes in asymptotically
%significant ways.

Another class of precomputation techniques is incremental
computation, where a program can efficiently respond to small changes
to its input by only recomputing parts of the computation affected
by the
changes~\cite{DemersReTe81,PughTe89,RamalingamRe93,AcarBlBlHaTa09}.
Like other precomputation techniques, incremental computation can be
thought of as a staged model, where changes in inputs come over a
period of time in stages.
%
However, incremental computation does not requiring fixing any of its
inputs and allows, in the general case, for all inputs to
change. Thus, its benefits are dependent on the input changes made and
techniques typically utilize significant run-time (dynamic)
information.  For example, while we can apply incremental computation
to our quickselect example (\ref{sec:overview}), doing so would not
realize the same bounds as splitting since incremental computation
techniques would unfold the quickselect function based on the demanded
ranks, which can require linear time (as opposed to logarithmic) at
each stage.  Such an implementation would also maintain sophisticated
data structures dynamically at run time, the benefits of which depend
highly on input changes applied.

\paragraph{Frequency-reduction via pass separation.}
Perhaps the closest work to ours is the algorithm for data
specialization given by Knoblock and Ruf~\cite{knoblock96}, which
also seeks to statically split an explicitly staged program into two
stages.  However, they only consider a simple first-order language and
straight-line programs. Their work also treats all computations
guarded by second-stage conditionals as second-stage computations,
which would prevent optimization (via splitting) of programs such as
quickselect.

Automatic pass separation has also been a topic of interest in
programming systems for computer graphics, where, to achieve
high performance, programs are typically manually separated into
components by frequency of execution (Indeed, separation into a pre-determined
set of per-frequency modules is \emph{required} by the execution model
of modern graphics architectures.)  The software engineering
challenges of modifying multiple per-stage programs have led to
suggestions of writing graphics programs in an explicitly-staged
programming language\,\cite{Proudfoot:2001,Foley:2011,He:2014}, and
deferring the task of pass separation to the compiler. However, all
prior efforts, like that of Knoblock and Ruf\,\cite{knoblock96} have
been limited to simple, imperative languages. We are not aware of
prior work that presents a splitting algorithm in the context of
higher-order functional languages.
